Lossy compression of DNA sequencing quality data

### 2015-04-25 manuscript/q-compression.tex
Numerous small changes, particularly on table and table header to make things more accurate and consistent.

\title[lossy-compression]{Lossy compression of DNA sequencing quality
  data} \author[El Hadidi \textit{et~al.}]{Mohamed El
  Hadidi\,$^{1}$\footnote{authors contributed equally}, Christopher
  M. Hill\,$^{2}$\footnotemark[1], Andr\'{a}s
  Szolek\,$^{3}$\footnotemark[1], and Michael
  P. Cummings\,$^4$\footnote{to whom correspondence should be
    addressed}} \address{$^{1}$Department of Algorithms in
  Bioinformatics, Center for Bioinformatics, University of
  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany
  \\ $^{2}$Department of Computer Science, University of Maryland,
  College Park, Maryland, 20742 USA\\ $^{3}$Department of Applied
  Bioinformatics, Center for Bioinformatics, Quantitative Biology
  Center, and Department of Computer Science, University of
  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\ $^{4}$Center
  for Bioinformatics and Computational Biology, University of
  Maryland, College Park, 20742 USA}
%$\dagger$authors contributed equally
\history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
% Hack to display authors contribution.
\editor{Associate Editor: XXXXXXX}

@@

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data.
Sickle starts at the ends of the sequence and uses a sliding window ($0.1 \times$ the read length) to find locations where the window's average quality falls below a given threshold (20 by default).
The sequence is then trimmed from this position to the respective end.
%uses a variable sliding window approach where base qualities are averaged over a window ($0.1 \times$ the read length) and trims from the position where the window's average quality falls below a given threshold (20 by default).
If the trimmed sequence's length is less than a certain threshold (20 by default), then the sequence is discarded.

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{MaxQual} & \multicolumn{2}{c|}{MinQual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{Degree 0} & \multicolumn{2}{c}{Degree 1} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{Degree 3} & \multicolumn{2}{c|}{Degree 5} & \multicolumn{2}{c|}{Degree 7} & \multicolumn{2}{c|}{Profile (64)} & \multicolumn{2}{c}{Profile (128)} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{Profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\

\end{tabular}

>>>>

\title[lossy-compression]{Lossy compression of DNA sequencing quality
  data} \author[El Hadidi \textit{et~al.}]{Mohamed El
  Hadidi\,$^{1}$\footnote{authors contributed equally}, Christopher
  M. Hill\,$^{2}$\footnotemark[1], Andr\'{a}s
  Szolek\,$^{3}$\footnotemark[1], and Michael
  P. Cummings\,$^4$\footnote{to whom correspondence should be
    addressed}} \address{$^{1}$Department of Algorithms in
  Bioinformatics, Center for Bioinformatics, University of
  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany
  \\ $^{2}$Department of Computer Science, University of Maryland,
  College Park, Maryland, 20742 USA\\ $^{3}$Department of Applied
  Bioinformatics, Center for Bioinformatics, Quantitative Biology
  Center, and Department of Computer Science, University of
  T\"{u}bingen, Sand 14, 72076 T\"{u}bingen, Germany\\ $^{4}$Center
  for Bioinformatics and Computational Biology, University of
  Maryland, College Park, Maryland, 20742 USA}
%$\dagger$authors contributed equally
\history{Received on XXXXX; revised on XXXXX; accepted on XXXX}
% Hack to display authors contribution.
\editor{Associate Editor: XXXXXXX}

@@

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the sequence and uses
a sliding window ($0.1 \times$ the read length) to find locations
where the window's average quality falls below a given threshold (20
by default). The sequence is then trimmed from this position to the
respective end.
%uses a variable sliding window approach where base qualities are
%averaged over a window ($0.1 \times$ the read length) and trims from
%the position where the window's average quality falls below a given
%threshold (20 by default).
If the trimmed sequence length is less than a certain threshold (20 by
default), then the sequence is discarded.

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Comparison of mapping for original reads and reads with
  compressed/decompressed quality values. Reads and reference genome
  are for \textit{Rhodobacter sphaeroides}, and mapping was performed
  using Bowtie2.}
\begin{small}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{max-qual} & \multicolumn{2}{c|}{min-qual} & \multicolumn{2}{c|}{2-bin} & \multicolumn{2}{c|}{regression (0)} & \multicolumn{2}{c}{regression (1)} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
& proportion & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{regression (3)} & \multicolumn{2}{c|}{regression (5)} & \multicolumn{2}{c|}{regression (7)} & \multicolumn{2}{c|}{profile (64)} & \multicolumn{2}{c}{profile (128)} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
& proportion & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{profile (256)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (6)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (10)} & \multicolumn{2}{c|}{\textsc{q}ual\textsc{c}omp (30)} & \multicolumn{2}{c}{\textsc{q}ual\textsc{c}omp (100)} \\
& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
& proportion & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
\end{tabular}
\end{small}
